---
title: "introduction-to-hood2net"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction-to-hood2net}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction and Motivation 

UNDER CONSTRUCTION

The `hood2net` package takes a list of words and/or their phonological transcriptions and creates a language network based on their neighborhood structure.

First, the phonological/orthographic neighbors for each item in the list are identified based on various definitions of a "neighbor". A pair of words can be considered to be neighbors (and thus become connected in the network) via the following ways: edit-distance (substitution, deletion, or addition; i.e., Levenshtein) or substitution only (i.e., Hamming). The default setting uses a distance of 1, but larger distances can be specified for a more liberal definition of a neighbor. The segmentation of the transcription can also be specified: either based on single character (i.e., single letter or phoneme) or user-specified segments indicated by separators (e.g., larger chunks like syllables or morphemes that are more than a single character separated by a '.'). 

`hood2net` then summarizes the neighborhood information for all items in the list into an `igraph` network object for subsequent analyses. Helper functions for extracting network metrics, neighborhood size, and other information from the language network are provided. This package is intended for psycholinguists interested in modeling language networks and lexical neighborhoods in various languages.


# Set up 

`hood2net` can be downloaded from Github while the package is being prepared for submission to CRAN as follows: 

```{r setup}
#| echo: true
#| eval: true
# install devtools first if needed
# devtools::install_github('csqsiew/hood2net')

# then load the package
library(hood2net)
```
# Creating networks

We'll create our first language network from a sample list of words provided in the package. Let's view the list:

```{r}
#| echo: true
#| eval: true
sample1
```

Notice that it is a list of words and the second column contains the number of letters. Let's say we want to create an orthographic similarity network where pairs of words that are different by the substitution, deletion or addition of one letter...

```{r}
#| echo: true
#| eval: true

g1 <- make_network(sample1)

library(igraph)

plot(g1)
```

# Computing network and neighborhood measures 

# Case Studies and Examples

## Case Study 1: 

## Case Study 2: 
