---
title: "A Quick Introduction to hood2net"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction-to-hood2net}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction and Motivation 

The `hood2net` package takes a list of words and/or their phonological transcriptions and creates a language network based on their phonological or orthographic neighborhood structure. This is where the name of the package comes from: From a word's neighborHOOD to a NETwork.

First, the phonological/orthographic neighbors for each item in the list are identified based on various definitions of a "neighbor". A pair of words can be considered to be neighbors (and thus become connected in the language network) via the following ways: edit-distance (substitution, deletion, or addition; i.e., Levenshtein distance) or substitution only (i.e., Hamming distance). The default setting uses a distance of 1, but larger distances can be specified for a more liberal definition of a neighbor. The segmentation of the transcription can also be specified: either based on single character (i.e., single letter or phoneme) or user-specified segments indicated by separators (e.g., larger chunks such as diphthongs, syllables, or morphemes that are longer than a single character separated by a symbol like '.' [period] or ' ' [space]). 

`hood2net` then summarizes the neighborhood information for all items in the list into an `igraph` network object for subsequent analyses. Helper functions for extracting network metrics, neighborhood size, and other information from the language network are provided. This package is intended for psycholinguists interested in modeling language networks and lexical neighborhoods in various languages.

## Why `hood2net`? 

Although there are several resources that exist for calculating the neighborhoods or similarity measures of words in language, to the best of my knowledge none of these existing resources provide capabilities to convert the neighborhood structure of words into a network, which is necessary for network analyses of such lexical information. Below are a few examples and a brief statement of their purpose and function and how it compares to `hood2net`: 

*Jiwar* is a multilingual neighborhood calculator and database for 40 languages of various scripts (Alzahrani, 2025). It provides phonological, orthographic, and phonographic neighborhood measures for these languages based on the OpenSubtitles corpus. Although it is a highly comprehensive database and neighborhood calculator, Jiwar does not provide the functions to convert the neighborhood information into a language network. The multilingual word lists and phonological transcriptions from Jiwar could be provided to `hood2net` to produce language networks and extract additional types of network metrics (e.g., closeness centrality or eigenvector centrality). It is also worth noting that the vast majority of neighborhood calculators or databases are limited to a single language; hence another strength of `hood2net` is the ability to compute neighborhood measures for languages not yet represented or unavailable in the literature. 

Another database worth mentioning is by Alderete et al. (2025), who created phonological similarity networks of English based on the SUBTLEX-US English corpus. Although this database contains phonological networks, it is limited to English. By allowing user-specified word lists, `hood2net` allows for word similarity networks to be constructed for a variety of languages, enabling network analyses to be extended to languages beyond English. 

# Set up 

`hood2net` can be downloaded from Github while the package is being prepared for submission to CRAN as follows. In order to compile the package on your local computer, Windows users will need to install... 

```{r setup}
#| echo: true
#| eval: true
# install devtools first if needed
# devtools::install_github('csqsiew/hood2net')

# then load the package
library(hood2net)
```

# Creating networks

We'll create our first language network from a sample list of words provided in the package. Let's view the list:

```{r}
#| echo: true
#| eval: true
sample1
```

Notice that it is a list of words and the second column contains the number of letters. Let's say we want to create an orthographic similarity network where pairs of words that are different by the substitution, deletion or addition of one letter. We can run the following R code as follows: 

```{r}
#| echo: true
#| eval: true

g1 <- make_network(sample1)

library(igraph)

plot(g1, vertex.frame.color = 'white', vertex.label.dist = 2.5,
     frame = TRUE, main = 'Single character, 1-edit Levenshtein distance')
```

Notice that the default settings of `make_network` is to segment each item based on single characters, and to create neighbors based on 1-edit (Levenshtein) distance. 

If we would like to create substitution-only networks, we can easily do this by changing the arguments in `make_network`: 

```{r}
#| echo: true
#| eval: true

g2 <- make_network(sample1, neighbor_type = 'hamming') # sub-only

plot(g2, vertex.frame.color = 'white', vertex.label.dist = 2.5,
     frame = TRUE, main = 'Single character, Hamming distance of 1')
```

For a more liberal operationalization of neighbors, we can also allow pairs of words to be neighbors if they are different by up to 2-edit distance, by adapting the code as follows: 

```{r}
#| echo: true
#| eval: true

g3 <- make_network(sample1, edit_size = 2)

plot(g3, vertex.frame.color = 'white', vertex.label.dist = 2.5,
     frame = TRUE, main = 'Single character, 2-edist Levenshtein distance')
```

Notice that in this network, the words 'heat' and 'hate' are connected as a deletion and an addition (2 operations) are needed to convert from one string into the other. 

If you don't want to use single characters as the segmentation scheme, the `make_network_sep` function can be used instead and each item can be segmented based on the separator symbol specified. In this sample list, we have the same list of items as before, but with periods (.) separating each letter. We can use the `make_network_sep` function to create a network by segmenting the items based on the periods: 

```{r}
#| echo: true
#| eval: true
sample2

g4 <- make_network_sep(sample2, separator = '.')

plot(g4, vertex.frame.color = 'white', vertex.label.dist = 2.5,
     frame = TRUE, main = 'Period separator, 1-edit Levenshtein distance')
```

This function may be useful for situations where you would like subgroups of characters to be considered as a single unit, for instance, if there are diphthongs (e.g., "\\aɪ\\") or if you wish to explore higher-order structures like syllabic structure. The default separator symbol is the period, but other symbols like ' ' [space] are also permitted. The other arguments mentioned, `edit_size` and `neighbor_type`, are also applicable for the `make_network_sep` function.

*Pro-tip*: You may want to save very large networks as they take a long time to complete. This can be done easily and then you can reload the file into your next R session. 

```{r}
#| echo: true
#| eval: false

# save
saveRDS(g1, file = 'my-network.RDS')

# load 
readRDS('my-network.RDS')
```

# Computing network and neighborhood measures 

Once the language network is generated, it can be used to (i) obtain global-level or macro-level information about the network structure, and (ii) generate neighborhood-metrics for individual words.

## Obtaining macro-level network measures 

The helper function `get_network_info` can be used to get this information readily: 

```{r}
#| echo: true
#| eval: true
get_network_info(g1)
```

It returns in a named vector: 

1. `nodes`: The number of nodes.
2. `edges`: The number of edges. 
3. `ASPL`: Average shortest path length refers to the mean of the shortest path between all possible node pairs (that have an non-infinite path).
4. `global CC`: Global clustering coefficient or transitivity refers to the proportion of closed triangles in the network relative to the number of possible triangles. It is a measure of overall level of local connectivity among nodes in the network.
5. `mean CC`: The average local clustering coefficient refers to the mean of each node's local clustering coefficient, which measures the amount of interconnectivity in the local neighborhood of each node. 
6. `diameter`: Diameter refers to the length of the longest short path between node pairs in the network. 
7. `mean degree`: The average degree of all nodes in the network. Degree refers to the number of links (or neighbors) that each node has. 
8. `no. components`: The number of distinct, separate components in the network. 
9. `prop. in LCC`: The proportion of nodes found in the largest connected component of the network. 

For more details about these measures and their application to language or cognitive networks, please refer to [https://csqsiew.github.io](https://csqsiew.github.io) (an online tutorial "Introduction to Network Analysis in `igraph`") and Siew et al. (2019). 

If you wish to compute other network measures, you can simply pass the network object into the relevant `igraph` function. Below we compute the density of the network `g1`:

```{r}
#| echo: true
#| eval: true

library(igraph)

edge_density(g1)
```

## Compute micro-level network measures for individual nodes 

Another cool feature of `hood2net` is the ability to easily extract the neighborhood information for all or subsets of nodes in the network. 

To obtain neighborhood size, which corresponds to node degree: 

```{r}
#| echo: true
#| eval: true

get_neighbor_size(g1)
```

To obtain local clustering coefficient, which is a measure of how interconnected a node's neighborhood is: 

```{r}
#| echo: true
#| eval: true

get_neighbor_clustering(g1)
```

It is also possible to compute the mean of a node's neighbors' property. Recall that in `sample1`, there was a second column depicting the length of the word: 

```{r}
#| echo: true
#| eval: true
sample1
```

In the constructed network, `length` is included as a node or vertex attribute. 

```{r}
#| echo: true
#| eval: true

summary(g1)
```

This makes it possible to obtain the mean length of each node's neighbor, as follows: 

```{r}
#| echo: true
#| eval: true

get_neighbor_mean(g1, attribute = 'length')
```

Notice that the name of the node attribute must be specified. A useful application of this function is that lexical properties of words can be appended to the list that was initially submitted to `make_network`, and automatically embedded into the network as node attributes. If word frequency were included, it would be possible use this function to compute mean neighborhood frequency for all items in the network, and also compute macro-level network information such as assortative mixing by frequency (i.e., do high-frequency words tend to be connected to other high-frequency words in the network?). 

# References 

Alderete, J., Mann, S., & Tupper, P. (2025). Open-access network science: Investigating phonological similarity networks based on the SUBTLEX-US lexicon. *Behavior Research Methods, 57*(3), 96. [https://doi.org/10.3758/s13428-025-02610-9](https://doi.org/10.3758/s13428-025-02610-9)

Alzahrani, A. (2025). Jiwar: A database and calculator for word neighborhood measures in 40 languages. *Behavior Research Methods, 57*(3), 98. [https://doi.org/10.3758/s13428-025-02612-7](https://doi.org/10.3758/s13428-025-02612-7)

Siew, C. S. Q., Wulff, D. U., Beckage, N. M., & Kenett, Y. N. (2019). Cognitive Network Science: A Review of Research on Cognition through the Lens of Network Representations, Processes, and Dynamics. *Complexity*, e2108423. [https://doi.org/10.1155/2019/2108423](https://doi.org/10.1155/2019/2108423)

